Adversarial Variational Optimization of\\ Non-Differentiable Simulators

### 2018-01-25 icml/avo.tex
init icml version

\documentclass[twoside]{article}
\usepackage{aistats2018}
 % If your paper is accepted, change the options for the package
% aistats2018 as follows:
%
%\usepackage[accepted]{aistats2018}
%
% This option will print headings for the title of your paper and
% headings for the authors names, plus a copyright note at the end of
% the first column of the first page.

@@

\usepackage{bm}
\usepackage{cancel}
\usepackage{bbold}
\usepackage{bm}
\usepackage{slashed}
\usepackage{graphicx}
\usepackage{color}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{tikz}

@@

@@

\aistatstitle{Adversarial Variational Optimization of\\
              Non-Differentiable Simulators}

\aistatsauthor{Author 1 \And Author 2}
\aistatsaddress{Institution 1 \And Institution 2}

% \aistatsauthor{Gilles Louppe \And Kyle Cranmer}
%
% \aistatsaddress{New York University\\
%                 \& University of Li{\`e}ge
%                 \And New York University}
                ]

@@

In this note, we develop a likelihood-free inference algorithm
for non-differentiable, implicit generative models.
We adapt the adversarial
training procedure of generative adversarial
networks~\cite{goodfellow2014generative} by replacing the implicit generative
network with a domain-based scientific simulator, and solve the resulting
non-differentiable minimax problem by minimizing variational upper
bounds~\cite{2011arXiv1106.4487W,2012arXiv1212.4507S} of the adversarial
objectives.
The objective of both algorithms is to match marginal distribution of
the generated data to the empirical distribution of the observations.

@@

Given some observed data $\{ \mathbf{x}_i | i=1, \dots, N \}$ drawn from the
(unknown) true distribution $p_r(\mathbf{x})$, our goal is to estimate the parameters
$\bftheta^*$ that minimize the divergence between $p_r(\mathbf{x})$ and
the implicit model $p(\mathbf{x}|\bftheta)$. That is,
\begin{equation}
    \bftheta^* = \arg \min_{\bftheta} \rho(p_r(\mathbf{x}), p(\mathbf{x}|\bftheta)),
\end{equation}
where $\rho$ is some distance or divergence.

@@

As thoroughly explored in \cite{2017arXiv170104862A}, GANs remain remarkably
difficult to train because of vanishing gradients as $d$ saturates, or because of
unreliable updates when the training procedure is relaxed. As a remedy,
Wasserstein GANs~\cite{2017arXiv170107875A} reformulate the adversarial
setup in order to minimize the Wasserstein-1 distance $W( p_r(\mathbf{x} ), p(\mathbf{x} | \bftheta)  )$
by replacing the adversarial classifier with a 1-Lipschitz adversarial critic
$d(\cdot; \bfphi) : \mathbb{R}^d \to \mathbb{R}$, such that
\begin{align}
& W( p_r(\mathbf{x} ), p(\mathbf{x} | \bftheta)  ) \nonumber \\
&= \sup_{\bfphi}  \mathbb{E}_{\tilde{\mathbf{x}}\sim p(\mathbf{x}|\bftheta)} [d(\tilde{\mathbf{x}}; \bfphi)] - \mathbb{E}_{\mathbf{x} \sim p_r(\mathbf{x})} [d(\mathbf{x}; \bfphi)] \nonumber \\
&\equiv \sup_{\bfphi} \mathcal{L}_W.
\end{align}
Under the WGAN-GP formulation of \cite{2017arXiv170400028G}
for stabilizing the optimization procedure,
training $d$ and $g$ results in alternating gradient updates on $\bfphi$ and $\bftheta$ in order to respectively minimize
\begin{align}
    {\cal L}_d =\,&  \mathcal{L}_W + \lambda \mathbb{E}_{\hat{\mathbf{x}} \sim p({\hat{\mathbf{x}})}} [(|| \nabla_{\hat{\mathbf{x}}} d({\hat{\mathbf{x}}};\bfphi) ||_2 - 1)^2] \\
    {\cal L}_g =\,& - \mathcal{L}_W
\end{align}
where ${\hat{\mathbf{x}}} := \epsilon \mathbf{x} +
(1-\epsilon)\tilde{\mathbf{x}}$, for $\epsilon \sim U[0,1]$, $\mathbf{x} \sim
p_r(\mathbf{x})$ and $\tilde{\mathbf{x}} \sim p(\mathbf{x}|\bftheta)$.

@@

Variational optimization~\cite{2012arXiv1212.4507S,staines2013optimization} and evolution strategies~\cite{2011arXiv1106.4487W} are general
optimization techniques that can be used to form a differentiable bound
on the optima of a non-differentiable function. Given a function $f$ to minimize,
these techniques are based on the simple fact that
\begin{equation}
    \min_{\bftheta} f(\bftheta) \leq \mathbb{E}_{\bftheta \sim q(\bftheta|\bfpsi)} [f(\bftheta)] = U(\bfpsi),
\end{equation}
where $q(\bftheta|\bfpsi)$ is a proposal distribution with parameters $\bfpsi$ over input values $\bftheta$.
That is, the minimum of a set of function values is always less than or equal
to any of their average. Provided that the proposal distribution is flexible enough, the parameters $\bfpsi$
can be updated to place its mass arbitrarily tight around the optimum $\bftheta^* = \min_{\bftheta \in \Theta} f(\bftheta)$.

@@

The alternating stochastic gradient descent on ${\cal L}_d$ and ${\cal L}_g$ in
GANs (Section~\ref{sec:gans}) inherently assumes that the generator $g$ is a differentiable function. In
the setting where we are interested in estimating the parameters of a
fixed non-differentiable simulator (Section~\ref{sec:problem}),
rather than in learning the generative model itself,
gradients $\nabla_\bftheta g$ either do not exist or are inaccessible. As a
result, gradients $\nabla_\bftheta {\cal L}_g$ cannot be constructed and the
optimization procedure cannot be carried out.

In this work, we propose to rely on variational optimization to minimize ${\cal
L}_d$ and ${\cal L}_g$, thereby bypassing the non-differentiability of $g$. More
specifically, we consider a proposal distribution $q(\bftheta|\bfpsi)$ over the
parameters of $g$ and alternately minimize the variational upper bounds
\begin{align}
U_d &= \mathbb{E}_{\bftheta \sim q(\bftheta|\bfpsi)} [ {\cal L}_d ] \label{eqn:vo-ud} \\
U_g &= \mathbb{E}_{\bftheta \sim q(\bftheta|\bfpsi)} [ {\cal L}_g ] \label{eqn:vo-ug}
\end{align} respectively over $\bfphi$ and $\bfpsi$.
When updating
$\bfphi$, unbiased estimates of $\nabla_\bfphi U_d$ can be obtained by
directly evaluating the gradient of $U_d$ over mini-batches of real and
generated data, as is ordinarily done in stochastic gradient descent. When updating
$\bfpsi$, $\nabla_\bfpsi U_g$ can be estimated as described in the previous section.
That is,
\begin{equation}\label{eqn:grad-ug-approx}
\nabla_\bfpsi U_g = \mathbb{E}_{\substack{\bftheta \sim q(\bftheta|\bfpsi), \\ \tilde{\bfx} \sim p(\bfx | \bftheta)}}  [-d( \tilde{\bfx} ;\bfphi) \nabla_\bfpsi \log q(\bftheta|\bfpsi)],
\end{equation}
which we can approximate with mini-batches of
generated data
\begin{equation}
\nabla_\bfpsi U_g \approx \frac{1}{M} \sum_{m=1}^M -d(g(\bfz_m; \bftheta_m); \bfphi) \nabla_\bfpsi \log q(\bftheta_m|\bfpsi)
\end{equation}
for $\bftheta_m \sim q(\bftheta|\bfpsi)$ and $\bfz_m \sim p(\bfz|\bftheta_m)$.
For completeness, Algorithm~\ref{alg:avo} outlines the proposed adversarial variational
optimization (AVO) procedure, as built on top of WGAN-GP.

Algorithm~\ref{alg:avo} represents the simplest version of AVO; however, the
variance of the noisy estimator of the gradients may be too large to be useful in many problems.
We use Adam~\cite{2014arXiv1412.6980K}, but note the opportunity to use instead the Natural Evolution Strategy
algorithm~\cite{2011arXiv1106.4487W} or variance reduction techniques
such as Rao-Blackwellization,
Control Variates, or quasi-Monte Carlo~\cite{ranganath2014black, tran2017variational}.

\subsection{Point Estimates through Marginal Likelihood}
%\subsection{Empirical Bayes through  Variational Inference}

@@

We can view the optimization of the marginal likelihood $q(\bfx | \bfpsi)$ with respect to $\bfpsi$ through the lens of empirical Bayes, where the data are used to optimize a prior within the family $q(\bftheta|\bfpsi)$. If $\rho$ is the KL distance, $\bfpsi^*$ would correspond to the maximum marginal likelihood estimator advocated by Rubin~\cite{rubin1984}.
When $\rho$ is the Wasserstein distance, $\bfpsi^*$ is referred to as the
minimum Wasserstein estimator (MWE).
When the model is well specified, the MWE
coincides with the true data-generating parameter; however, if the model is
misspecified, the MWE is typically different from the maximum likelihood
estimator (MLE).
Thus, if the simulator $p(\bfx |
\bftheta)$ is misspecified, $q(\bftheta | \bfpsi)$ will attempt to smear it so that the marginal
model $q(\bfx | \bfpsi)$ is closer to $p_r(\bfx)$. However, if the simulator is well specified,
then $q(\bftheta | \bfpsi)$ will concentrate its mass around the true data-generating parameter.

In order to more effectively target
point estimates $\bftheta^*$, we augment Eqn.~\ref{eqn:vo-ug} with a
regularization term corresponding to the differential entropy $H$ of the
proposal distribution $q(\bftheta | \bfpsi)$. That is,
\begin{equation}
U_g = \mathbb{E}_{\bftheta \sim q(\bftheta|\bfpsi)} [ {\cal L}_g ] + \gamma H(q(\bftheta|\bfpsi))
\end{equation}
where $\gamma \in \mathbb{R}^+$ is a hyper-parameter controlling the trade-off
between the generator objective and the tightness of the proposal distribution.

For small values of $\gamma$,
proposal distributions with large entropy are not penalized, which results
in learning a smeared variation of the original simulator corresponding to the marginal or prior predictive model. On the other hand,
for large values of $\gamma$, the procedure is encouraged to fit a proposal
distribution with low entropy, which has the effect of concentrating its density
tightly around one or a few $\bftheta$ values.

@@

\bibliographystyle{hunsrt}
\bibliography{bibliography.bib}

>>>>

\documentclass{article}
\usepackage[accepted]{icml2018}
 % If your paper is accepted, change the options for the package
% aistats2018 as follows:
%
%\usepackage[accepted]{aistats2018}
%
% This option will print headings for the title of your paper and
% headings for the authors names, plus a copyright note at the end of
% the first column of the first page.

@@

\usepackage{bm}
\usepackage{cancel}
\usepackage{bbold}
\usepackage{bm}
\usepackage{slashed}
\usepackage{graphicx}
\usepackage{color}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{tikz}
\usepackage{hyperref}

% Attempt to make hyperref and algorithmic work together better:
\newcommand{\theHalgorithm}{\arabic{algorithm}}

@@

\icmltitlerunning{Adversarial Variational Optimization of Non-Differentiable Simulators}

@@

\icmltitle{Adversarial Variational Optimization of\\
           Non-Differentiable Simulators}

\begin{icmlauthorlist}
\icmlauthor{Glles Louppe}{uliege}
\icmlauthor{Kyle Cranmer}{nyu}
\end{icmlauthorlist}

\icmlaffiliation{uliege}{University of Li{\`e}ge, Belgium}
\icmlaffiliation{nyu}{New York University, USA}

\icmlcorrespondingauthor{Gilles Louppe}{g.louppe@uliege.be}
\icmlcorrespondingauthor{Kyle Cranmer}{kyle.cranmer@nyu.edu}

% You may provide any keywords that you
% find helpful for describing your paper; these are used to populate
% the "keywords" metadata in the PDF but will not be shown in the document
\icmlkeywords{XXXXX}

\vskip 0.3in
]

\printAffiliationsAndNotice{\icmlEqualContribution}

@@

In this note, we develop a likelihood-free inference algorithm
for non-differentiable, implicit generative models.
We adapt the adversarial
training procedure of generative adversarial
networks~\cite{goodfellow2014generative} by replacing the implicit generative
network with a domain-based scientific simulator, and solve the resulting
non-differentiable minimax problem by minimizing variational upper
bounds~\cite{2011arXiv1106.4487W,2012arXiv1212.4507S} of the adversarial
objectives.
The objective of the algorithm is to match the marginal distribution of
the generated data to the empirical distribution of the observations.

@@

Given some observed data $\{ \mathbf{x}_i | i=1, \dots, N \}$ drawn from the
(unknown) true distribution $p_r(\mathbf{x})$, our goal is to estimate the parameters
$\bftheta^*$ that minimize the divergence or the distance $\rho$ between $p_r(\mathbf{x})$ and
the implicit model $p(\mathbf{x}|\bftheta)$. That is,
\begin{equation}
    \bftheta^* = \arg \min_{\bftheta} \rho(p_r(\mathbf{x}), p(\mathbf{x}|\bftheta)).
\end{equation}

@@

As thoroughly explored in \cite{2017arXiv170104862A}, GANs remain remarkably
difficult to train because of vanishing gradients as $d$ saturates, or because of
unreliable updates when the training procedure is relaxed. As a remedy,
Wasserstein GANs~\cite{2017arXiv170107875A} (WGANs) reformulate the adversarial
setup in order to minimize the Wasserstein-1 distance $W( p_r(\mathbf{x} ), p(\mathbf{x} | \bftheta)  )$
by replacing the adversarial classifier with a 1-Lipschitz adversarial critic
$d(\cdot; \bfphi) : \mathbb{R}^d \to \mathbb{R}$, such that
\begin{align}
& W( p_r(\mathbf{x} ), p(\mathbf{x} | \bftheta)  ) \nonumber \\
&= \sup_{\bfphi}  \mathbb{E}_{\tilde{\mathbf{x}}\sim p(\mathbf{x}|\bftheta)} [d(\tilde{\mathbf{x}}; \bfphi)] - \mathbb{E}_{\mathbf{x} \sim p_r(\mathbf{x})} [d(\mathbf{x}; \bfphi)] \nonumber \\
&\triangleq \sup_{\bfphi} \mathcal{L}_W.
\end{align}
Under the WGAN-GP formulation of \cite{2017arXiv170400028G}
for stabilizing the optimization procedure,
training $d$ and $g$ results in alternating gradient updates on $\bfphi$ and $\bftheta$ in order to respectively minimize
\begin{align}
    {\cal L}_d =\,&  \mathcal{L}_W + \lambda \mathbb{E}_{\hat{\mathbf{x}} \sim p({\hat{\mathbf{x}})}} [(|| \nabla_{\hat{\mathbf{x}}} d({\hat{\mathbf{x}}};\bfphi) ||_2 - 1)^2] \\
    {\cal L}_g =\,& - \mathcal{L}_W
\end{align}
where ${\hat{\mathbf{x}}} := \epsilon \mathbf{x} +
(1-\epsilon)\tilde{\mathbf{x}}$, for $\epsilon \sim U[0,1]$, $\mathbf{x} \sim
p_r(\mathbf{x})$ and $\tilde{\mathbf{x}} \sim p(\mathbf{x}|\bftheta)$.

@@

Variational optimization~\cite{2012arXiv1212.4507S,staines2013optimization} and evolution strategies~\cite{2011arXiv1106.4487W} are general
optimization techniques that can be used to form a differentiable bound
on the optima of a non-differentiable function. Given a function $f$ to minimize,
these techniques are based on the observation that
\begin{equation}
    \min_{\bftheta} f(\bftheta) \leq \mathbb{E}_{\bftheta \sim q(\bftheta|\bfpsi)} [f(\bftheta)] = U(\bfpsi),
\end{equation}
where $q(\bftheta|\bfpsi)$ is a proposal distribution with parameters $\bfpsi$ over input values $\bftheta$.
That is, the minimum of a set of function values is always less than or equal
to any of their average. Provided that the proposal distribution is flexible enough, the parameters $\bfpsi$
can be updated to place its mass arbitrarily tight around the optimum $\bftheta^* = \min_{\bftheta \in \Theta} f(\bftheta)$.

@@

The alternating stochastic gradient descent on ${\cal L}_d$ and ${\cal L}_g$ in
WGANs (Section~\ref{sec:gans}) inherently assumes that the generator $g$ is a differentiable function. In
the setting where we are interested in estimating the parameters of a
fixed non-differentiable simulator (Section~\ref{sec:problem}),
rather than in learning the generative model itself,
gradients $\nabla_\bftheta g$ either do not exist or are inaccessible. As a
result, gradients $\nabla_\bftheta {\cal L}_g$ cannot be constructed and the
optimization procedure cannot be carried out.

In this work, we propose to rely on variational optimization to minimize ${\cal
L}_d$ and ${\cal L}_g$, thereby bypassing the non-differentiability of $g$. More
specifically, we consider a proposal distribution $q(\bftheta|\bfpsi)$ over the
parameters of the simulator $g$ and alternately minimize the variational upper bounds
\begin{align}
U_d &= \mathbb{E}_{\bftheta \sim q(\bftheta|\bfpsi)} [ {\cal L}_d ] \label{eqn:vo-ud} \\
U_g &= \mathbb{E}_{\bftheta \sim q(\bftheta|\bfpsi)} [ {\cal L}_g ] \label{eqn:vo-ug}
\end{align} respectively over $\bfphi$ and $\bfpsi$.
When updating
$\bfphi$, unbiased estimates of $\nabla_\bfphi U_d$ can be obtained by
directly evaluating the gradient of $U_d$ over mini-batches of real and
generated data. When updating
$\bfpsi$, $\nabla_\bfpsi U_g$ can be estimated as described in the previous section.
That is,
\begin{equation}\label{eqn:grad-ug-approx}
\nabla_\bfpsi U_g = \mathbb{E}_{\substack{\bftheta \sim q(\bftheta|\bfpsi), \\ \tilde{\bfx} \sim p(\bfx | \bftheta)}}  [-d( \tilde{\bfx} ;\bfphi) \nabla_\bfpsi \log q(\bftheta|\bfpsi)],
\end{equation}
which we can approximate with mini-batches of
generated data
\begin{equation}
\nabla_\bfpsi U_g \approx \frac{1}{M} \sum_{m=1}^M -d(g(\bfz_m; \bftheta_m); \bfphi) \nabla_\bfpsi \log q(\bftheta_m|\bfpsi)
\end{equation}
for $\bftheta_m \sim q(\bftheta|\bfpsi)$ and $\bfz_m \sim p(\bfz|\bftheta_m)$.
For completeness, Algorithm~\ref{alg:avo} outlines the proposed adversarial variational
optimization (AVO) procedure, as built on top of WGAN-GP.

Algorithm~\ref{alg:avo} represents the simplest version of AVO; however, the
variance of the noisy estimator of the gradients may be too large to be useful in many problems.
We use Adam~\cite{2014arXiv1412.6980K}, but note the opportunity to use instead the Natural Evolution Strategy
algorithm~\cite{2011arXiv1106.4487W} or variance reduction techniques
such as Rao-Blackwellization,
control variates, or quasi-Monte Carlo~\cite{ranganath2014black, tran2017variational}.

\subsection{Parameter Point Estimates}
%\subsection{Empirical Bayes through  Variational Inference}

@@

We can view the optimization of $q(\bfx | \bfpsi)$ with respect to $\bfpsi$ through the lens of empirical Bayes, where the data are used to optimize a prior within the family $q(\bftheta|\bfpsi)$. If $\rho$ is the KL distance, $\bfpsi^*$ would correspond to the maximum marginal likelihood estimator advocated by Rubin~\cite{rubin1984}.
When $\rho$ is the Wasserstein distance, $\bfpsi^*$ is referred to as the
minimum Wasserstein estimator (MWE).
When the model is well specified, the MWE
coincides with the true data-generating parameter; however, if the model is
misspecified, the MWE is typically different from the maximum likelihood
estimator (MLE).
Thus, if the simulator $p(\bfx |
\bftheta)$ is misspecified, $q(\bftheta | \bfpsi)$ will attempt to smear it so that the marginal
model $q(\bfx | \bfpsi)$ is closer to $p_r(\bfx)$. However, if the simulator is well specified,
then $q(\bftheta | \bfpsi)$ will concentrate its mass around the true data-generating parameter.

In order to more effectively target
point estimates $\bftheta^*$, we augment Eqn.~\ref{eqn:vo-ug} with
an entropic regularization term $H(q(\bftheta|\bfpsi))$, that is
\begin{equation}
U_g = \mathbb{E}_{\bftheta \sim q(\bftheta|\bfpsi)} [ {\cal L}_g ] + \gamma H(q(\bftheta|\bfpsi))
\end{equation}
where $\gamma \in \mathbb{R}^+$ is a hyper-parameter controlling the trade-off
between the generator objective and the tightness of the proposal distribution.
For small values of $\gamma$,
proposal distributions with large entropy are not penalized, which results
in learning a smeared variation of the original simulator corresponding to the marginal or prior predictive model. On the other hand,
for large values of $\gamma$, the procedure is encouraged to fit a proposal
distribution with low entropy, which has the effect of concentrating its density
tightly around one or a few $\bftheta$ values.

@@

\bibliography{bibliography.bib}
\bibliographystyle{icml2018}

\appendix

\section{Further experiments}
